---
title: "Buffalo Creek Flood Risk Assessment"
author: "Sazzad Hossain"
subtitle: "Find whether any infrastructure is at risk of flooding"
date: 2025-11-28
date-format: long
code-fold: true
code-tools:
  toggle: true
format:
  html:
    theme: default
    toc: true
    toc-depth: 3
    self-contained: true
---

# Introduction

Buffalo Creek, located in Erie County, New York, has a history of flooding that can threaten local infrastructure.  
This project aims to identify areas along Buffalo Creek that may put buildings or other structures at risk of flooding.  

The analysis combines publicly available elevation data, hydrological flowlines, and OpenStreetMap building data to generate spatial visualizations of flood buffers and at-risk infrastructure.  

The main questions addressed are:

1. Which buildings fall within potential flood zones of Buffalo Creek?
2. How many buildings are at risk within a 500-meter buffer of the stream?
3. Can this workflow be reproduced using publicly available data and code?

---

# Data

The analysis uses the following sources:

- **Digital Elevation Model (DEM):** downloaded automatically from AWS terrain using the `elevatr` package. Provides elevation information to help understand terrain and flood flow.
- **Hydrological data:** downloaded from NHDPlus High-Resolution using `nhdplusTools`. Contains stream and river networks.
- **Buildings:** downloaded from OpenStreetMap using `osmdata`. Used to identify structures potentially at risk.



# Load Libraries

These libraries allow downloading, processing, and visualizing both raster and vector spatial data, and creating figures and tables for the HTML report.



```{r message=FALSE, warning=FALSE}
library(nhdplusTools)
library(sf)
library(terra)
library(elevatr)
library(dataRetrieval)
library(osmdata)
library(ggplot2)
library(dplyr)
library(leaflet)
library(htmlwidgets)
library(knitr)
library(tmap)
```


## Digital Elevation Model (DEM)

We define a rectangular study area for Buffalo Creek.

DEM is downloaded automatically from AWS using elevatr.

This raster shows elevation values which help understand which areas are prone to flooding.

```{r message=FALSE, warning=FALSE}
#Define Buffalo Creek study area
bbox_coords <- c(xmin = -78.85, ymin = 42.80, xmax = -78.65, ymax = 42.92)
study_bbox <- st_bbox(bbox_coords, crs = 4326)
study_area <- st_as_sfc(study_bbox) |> st_sf()

#Download DEM
dem_raw <- get_elev_raster(locations = study_area, z = 13, src = "aws")
dem <- rast(dem_raw)
writeRaster(dem, "data/buffalo_dem.tif", overwrite = TRUE)

print(paste("DEM Resolution:", res(dem), "meters"))
rng <- minmax(dem)
print(paste("Elevation range:", round(rng[1]), "to", round(rng[2]), "meters"))

#Visualize DEM

plot(dem)
```

## Stream Extraction

Hydrological flowlines are downloaded to identify the stream network.

Only flowlines containing "Buffalo" are selected.

This is the basis for creating flood buffers.

```{r message=FALSE, warning=FALSE}
library(nhdplusTools)
library(sf)

# Define area of interest as sf polygon (not bbox)

aoi <- st_as_sfc(st_bbox(c(
  xmin = -78.9,
  ymin = 42.75,
  xmax = -78.6,
  ymax = 42.95
), crs = 4326))



# Download flowlines for the area
flowlines <- get_3dhp(AOI = aoi, type = "flowline")

# Filter for Buffalo Creek
buffalo_creek <- flowlines |>
  filter(grepl("Buffalo", gnisidlabel, ignore.case = TRUE))

# Visualize just the stream
plot(st_geometry(buffalo_creek), col = "blue", lwd = 2, main = "Buffalo Creek Stream Network")
```


## Flood Buffer Creation

We create a 500-meter buffer around Buffalo Creek to represent potential flood risk areas.

Buffer is transformed back to WGS84 for mapping and spatial intersection with buildings.

```{r message=FALSE, warning=FALSE}
buffalo_creek_proj <- st_transform(buffalo_creek, crs = 32617) # UTM zone for your area
stream_buffer <- st_buffer(buffalo_creek_proj, dist = 500) # 500 meters

# Transform back to WGS84 for plotting or OSM intersection
stream_buffer_wgs84 <- st_transform(stream_buffer, crs = 4326)

# Plot buffer and stream
plot(st_geometry(stream_buffer_wgs84), col = rgb(1,0,0,0.2), main = "Buffalo Creek 500m Flood Buffer")
plot(st_geometry(buffalo_creek), col = "blue", lwd = 2, add = TRUE)


```

## Building Data from OpenStreetMap

Building footprints are downloaded and transformed to match buffer CRS.

Intersection identifies buildings inside the 500-meter flood buffer.

```{r message=FALSE, warning=FALSE}
# Define the study area polygon for OSM query
bbox <- st_bbox(stream_buffer_wgs84)

# Get buildings from OSM
buildings_osm <- opq(bbox = bbox) %>%
  add_osm_feature(key = "building") %>%
  osmdata_sf()

# Extract building polygons
buildings <- buildings_osm$osm_polygons

# Ensure same CRS for intersection
buildings <- st_transform(buildings, st_crs(stream_buffer_wgs84))
```

## Identify Buildings at Risk

```{r message=FALSE, warning=FALSE}
buildings_at_risk <- st_intersection(buildings, stream_buffer_wgs84)

# Check how many buildings are affected
cat("Number of buildings at risk:", nrow(buildings_at_risk), "\n")

```
## Visualization
The map shows flood buffer in red, Buffalo Creek in blue, all buildings in grey, and buildings at risk in orange.

This provides a clear visualization of potential risk areas.

```{r message=FALSE, warning=FALSE}
library(tmap)

# Static plot for Quarto rendering
tmap_mode("plot")

tm_shape(stream_buffer_wgs84) +
  tm_fill(col = "red", alpha = 0.2) +
  tm_shape(buffalo_creek) +
  tm_lines(col = "blue") +
  tm_shape(buildings) +
  tm_polygons(col = "grey", alpha = 0.5) +
  tm_shape(buildings_at_risk) +
  tm_polygons(col = "orange", alpha = 0.8)


```
## Summary Table and Plot


```{r message=FALSE, warning=FALSE}
library(ggplot2)

# Total number of buildings
total_buildings <- nrow(buildings)

# Number of buildings at risk
at_risk <- nrow(buildings_at_risk)

# Number of buildings not at risk
not_at_risk <- total_buildings - at_risk

# Create summary data frame
summary_df <- data.frame(
  Status = c("At Risk", "Not at Risk"),
  Count = c(at_risk, not_at_risk)
)

# Print summary table
knitr::kable(summary_df, caption = "Summary of Buildings at Risk")

# Plot the summary
ggplot(summary_df, aes(x = Status, y = Count, fill = Status)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = Count), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("orange", "grey")) +
  labs(title = "Buildings at Risk vs Not at Risk",
       x = "", y = "Number of Buildings") +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(size = 14))

```

Conclusions

A 500-meter buffer around Buffalo Creek identifies buildings potentially exposed to flooding.

Future work could include interactive maps, multiple flood return periods, and more detailed infrastructure data.
